---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data

Setup environment in English
```{r setup, echo=TRUE}
Sys.setlocale("LC_TIME", "C")
```

### 1. Load the data

```{r load_data, echo=TRUE}
activity_data = read.csv("activity.csv", header=TRUE)
```

### 2. Check data structure

The variables included in this dataset are:

- **steps**: Number of steps taking in a 5-minute interval (missing values are coded as *NA*)
- **date**: The date on which the measurement was taken in YYYY-MM-DD format
- **interval**: Identifier for the 5-minute interval in which measurement was taken

```{r names_data, echo=TRUE}
names(activity_data)
```

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

```{r structure_data, echo=TRUE}
str(activity_data)
head(activity_data)
```


### 3. Process/transform the data

Convert the date column to Date format

```{r process_data, echo=TRUE}
activity_data$date = as.Date(activity_data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

### Calculate the total number of steps taken per day

Summarize the data (i.e. calculate the number of steps per day)

```{r summarize_data, echo=TRUE}
activity_data_summary = aggregate(activity_data["steps"], by=activity_data["date"], FUN=sum, na.rm=TRUE)
```

### Make a histogram of the total number of steps taken each day

```{r histogram, echo=TRUE}
hist(activity_data_summary$steps, 
     main="Histogram steps per day", 
     xlab="Number of Steps", 
     border="white", 
     col="grey",
     breaks=10)
```

### Calculate and report the mean and median of the total number of steps taken per day

Mean steps per day
```{r mean, echo=TRUE}
mean_steps = mean(activity_data_summary$steps)
hist(activity_data_summary$steps, 
     main="Histogram steps per day", 
     xlab="Number of Steps", 
     border="white", 
     col="grey",
     breaks=10)
abline(v=mean_steps, col="blue")
```

Median steps per day
```{r median, echo=TRUE}
median_steps = median(activity_data_summary$steps)
hist(activity_data_summary$steps, 
     main="Histogram steps per day", 
     xlab="Number of Steps", 
     border="white", 
     col="grey",
     breaks=10)
abline(v=median_steps, col="red")
```


## What is the average daily activity pattern?

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r daily_pattern, echo=TRUE}
daily_pattern = aggregate(activity_data["steps"], by=activity_data["interval"], FUN=mean, na.rm=TRUE)
plot(daily_pattern$interval, daily_pattern$steps, type="l", xlab="5-min Interval", ylab="Mean Steps")
```

5-minute interval which contains maximum number of steps

```{r max_avg_steps_interval, echo=TRUE}
max_mean_interval = max(daily_pattern$steps); max_mean_interval
max_interval = daily_pattern[which(daily_pattern$steps == max_mean_interval), ]$interval
plot(daily_pattern$interval, daily_pattern$steps, type="l", xlab="5-min Interval", ylab="Mean Steps")
abline(v=max_interval, col="red")
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset

```{r nas, echo=TRUE}
sum(is.na(activity_data$steps))
```

### Devise a strategy for filling in all of the missing values in the dataset and create a new dataset that is equal to the original dataset but with the missing data filled in
```{r imputting, echo=TRUE}
activity_data = transform(activity_data, steps = ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))
activity_data_summary = aggregate(activity_data["steps"], by=activity_data["date"], FUN=sum, na.rm=TRUE)
hist(activity_data_summary$steps, 
     main="Histogram steps per day", 
     xlab="Number of Steps", 
     border="white", 
     col="grey",
     breaks=10)
```

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekdays, echo=TRUE}
activity_data$dayofweek = weekdays(activity_data$date)
isWeekend = (activity_data$dayofweek %in% c("Saturday","Sunday"))
activity_data$typeofday = factor (isWeekend, labels = c("weekday","weekend"))
daily_pattern_2 = aggregate(activity_data["steps"], by = activity_data[c("interval", "typeofday")], FUN=mean, na.rm=TRUE)
str(daily_pattern_2)
```

### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r panelplot, echo=TRUE}
library(lattice)
xyplot(steps ~interval | typeofday, data = daily_pattern_2,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Mean Steps',
      layout = c(1,2))
```

